Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-02-03 paper/zhang-cook-laa-langrene-menendez.Rmd
update zhang-cook-laa-langrene-menendez.Rmd
minor corrections/typos

Several R implementations address optimization problems with both general purpose as well as task specifics solvers. The most prominent one within the general solvers is \code{optim()} in the \CRANpkg{stats} [@stats] package, which provides both gradient-based and derivative-free optimisation functions. Another general solver specialised in non-linear optimisation is \CRANpkg{nloptr} [@nloptr]. Specific solvers for simulated annealing includes \code{optim(..., method  "SANN")} and package \CRANpkg{GenSA} [@gensa] that deals with more complicated objective functions. For other task specific solvers, readers are recommended to visit the relevant sections in CRAN task review on \ctv{optimisation and mathematical programming} [@crantaskreviewoptim].

@@

Projection pursuit guided tour combines two different methods (projection pursuit and guided tour) in exploratory data analysis. Projection pursuit, coined by @friedman1974projection, detects interesting structures (e.g. clustering, outliers and skewness) in multivariate data via low dimensions projection. Guided tour is one variation of a broader class of data visualisation methods, tour, which displays high-dimensional data through a series of animated projections.

<!-- Notation is first established before defining projection pursuit. -->
Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A d-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto an scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. A number of index functions have been proposed in the literature to detect different data structures [friedman1974projection; @hall1989polynomial; @cook1993projection; @posse1995projection] and to classify groups [@lee2005projection; @lee2010projection].

As a general visualisation method, tour produces animations of high dimensional data via rotations of low dimension planes. There are different tour implementations depending on how the high dimensional space is investigated: grand tour [@cook2008grand] selects the planes randomly to provide a general overview;  manual tour [@cook1997manual] gradually phases in and out one variable to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses the planes with the aid of projection pursuit, to gradually reveal the most interesting projection. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs a geodesic interpolation between these planes to form a tour path. Figure \ref{fig:tour-path} shows a sketch of the tour path where the blue frames are produced by the projection pursuit optimisation and the white frames are interpolations  between the blue frames. Mathematical details of the geodesic interpolation can be found in @buja2005computational. The aforementioned tour method has been implemented in the R package \CRANpkg{tourr} [@tourr].

@@

The optimisation problem in projection pursuit is stated as follows: Given a randomly start basis $\mathbf{A}_1$, projection pursuit finds the final projection basis $\mathbf{A}_T$ that satisfies the following optimisation problem:

@@

\noindent where $I_d$ is the $d$-dimensional identity matrix and the constraint requires the projection bases $\mathbf{A}$ to be orthonormal. Several features of this optimisation are worth-noticing. 1) The optimisation is constrained and the space of the bases is a d-dimensional sphere. 2) The objective function may not be differentiable for a constructed index function. 3) Although finding the global maximum is the goal of an optimisation problem, it is also interesting to inspect local maximum in projection pursuit since it could present unexpected interesting projections. 4) The optimisation procedure needs to be easy to compute since the tour animation needs to be played in real-time during the optimisation.

@@

\begin{algorithm}
\SetAlgoLined
\Repeat{$l > l_{\max}$ or $p_{\text{diff}} > 0.001$}{
  generate $n$ random directions $\mathbf{A}_{\text{rand}}$ \;
  compute $2n$ candidate bases deviate from $\mathbf{A}_{\text{cur}}$ by an angle of $\delta$ while ensure orthogonality\;
  compute the corresponding index value for each candidate bases\;
  determine the search direction as from $\mathbf{A}_{\text{cur}}$ to the candidate bases with the largest index value\;
  determine the step size via optimising the index value on the search direction over a 90 degree window\;
  find the optima $\mathbf{A}_{**}$ and compute $I_{**} = f(\mathbf{A}_{**})$, $p_{\text{diff}} = (I_{**} - I_{\text{cur}})/I_{**}$\;
  update $l = l + 1$\;
}
\caption{Pseudo derivative (PD)}
\label{search-geodesic}
\end{algorithm}

@@

Three main pieces of information are recorded for the projection pursuit optimisers: 1) projection bases $\mathbf{A}$, 2) index values $I$, and 3) state $S$. For SA and SAJO, possible state include `random_search`,  `new_basis`, and `interpolation`. Pseudo derivative (PD) has a wider variety of state including `new_basis`, `direction_search`, `best_direction_search`, `best_line_search`, and `interpolation`. Multiple iterators index the information collected at different levels: $t$ is a unique identifier prescribing the natural ordering of each observation;  $j$ and $l$ are the counter of the outer and inner loop respectively. Other parameters of interest recorded include `method` that tags the name of the optimiser, and `alpha` that indicates the sampling neighbourhood size for searching observations.  A matrix notation of the data structure is presented in Equation \ref{eq:data-structure}.

@@

\noindent where $T^{\prime} = T + k_{J}+ l_{J+1}$. Note that there is no output in iteration $J + 1$ since the optimiser doesn't find a better basis in the last iteration and terminates. The final basis found is $A_T$ with index value $I_T$.

@@

A starting point of diagnosing an optimiser is to understand how many search an optimiser has conducted. One may want to simply plot the index value of the search points across its natural order, but a point geometry may work well if each iteration has a similar number of points. When some iterations have considerably more points than others, using a point geometry over-emphasizes the iterations that have more search points since these iterations will occupy the mast majority of the plot space. An alternative is to summarise the search in each iteration using boxplot and each iteration will then be spaced out equally. Occasionally, one may still want to switch back to a point geometry if the number of points is small in a particular iteration and this can be achieved via the `cutoff` argument in the search plot function. Additional annotations are added to facilitate better reading of the plot and these includes  
1) the number of points searched in each iteration can be added as text label at the bottom of each iteration; 2) the anchor bases to interpolate are connected and highlighted in a larger size; and 3) the colour of the last iteration is in a grey scale to indicate no better basis found in this iteration.

@@

```{r toy-search, fig.cap="A comparison of search by two otimisers: SA (left) and SAJO (right) on a 2D projection of a six-variable dataset, \\code{boa6}, using the holes index. Both optimisers find the final basis with similar index value while it takes SAJO more iteration because the algorithm allows a probabilistic acceptance of bases with lower index value, as observed in iteration 4 and 6-8."}
```

@@

```{r toy-interp, fig.cap = "Trace plots of the interpolating bases with the same optimisation routine as the previous figure. Both traces are smooth while the change of index value in each iteration may not be monotinic."}
```

Another interest of diagnostic is to examine how the index value improves for interpolating bases since the projection on these bases is played by the tour animation. Trace plots are created with plotting the index value of the interpolation points against time. Figure \ref{fig:toy-interp} presents the trace plot of the same optimisation routine as Figure \ref{fig:toy-search}. The trace plot is smooth and an interesting discovery is that in SAJO, the interpolation in iteration 6-8 is not monotonic. Sometimes, this non-monotonicity is acceptable since the target basis does not necessarily have a higher index value than the current one, while in some cases, the optimisation will be better off if these intermediate basis can be used as the start of the next iteration (see section \ref{monotonic} for a detailed discussion on this).

>>>>

Several R implementations address optimization problems with both general purpose as well as task specifics solvers. The most prominent one within the general solvers is \code{optim()} in the \CRANpkg{stats} [@stats] package, which provides both gradient-based and derivative-free optimisation functions. Another general solver specialised in non-linear optimisation is \CRANpkg{nloptr} [@nloptr]. Specific solvers for simulated annealing include \code{optim(..., method  "SANN")} and package \CRANpkg{GenSA} [@gensa] that deals with more complicated objective functions. For other task specific solvers, readers are recommended to visit the relevant sections in CRAN task review on \ctv{optimisation and mathematical programming} [@crantaskreviewoptim].

@@

Projection pursuit guided tour combines two different methods (projection pursuit and guided tour) in exploratory data analysis. Projection pursuit, coined by @friedman1974projection, detects interesting structures (e.g. clustering, outliers and skewness) in multivariate data via low-dimensional projections. Guided tour is one variation of a broader class of data visualisation methods, tour, which displays high-dimensional data through a series of animated projections.

<!-- Notation is first established before defining projection pursuit. -->
Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A $d$-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto a scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. A number of index functions have been proposed in the literature to detect different data structures [@friedman1974projection; @hall1989polynomial; @cook1993projection; @posse1995projection] and to classify groups [@lee2005projection; @lee2010projection].

As a general visualisation method, tour produces animations of high dimensional data via rotations of low-dimensional planes. There are different tour implementations depending on how the high dimensional space is investigated: grand tour [@cook2008grand] selects the planes randomly to provide a general overview;  manual tour [@cook1997manual] gradually phases in and out one variable to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses the planes with the aid of projection pursuit, to gradually reveal the most interesting projection. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs a geodesic interpolation between these planes to form a tour path. Figure \ref{fig:tour-path} shows a sketch of the tour path where the blue frames are produced by the projection pursuit optimisation and the white frames are interpolations  between the blue frames. Mathematical details of the geodesic interpolation can be found in @buja2005computational. The aforementioned tour method has been implemented in the R package \CRANpkg{tourr} [@tourr].

@@

In projection pursuit the optimisation problem \textcolor{red}{(aims at ... describe it in simple words.)} That is, it starts with a given randomly selected basis $\mathbf{A}_1$ and aims at finding an optimal final projection basis $\mathbf{A}_T$ in each iteration that satisfies the following optimisation problem:

@@

\noindent where $f$, $\mathcal{A}$, and $\mathbf{X}$ are defined as in section \textcolor{red}{XX}, and  $I_d$ is the $d$-dimensional identity matrix and the constraint requires the projection bases $\mathbf{A}$ to be orthonormal. Several features of this optimisation are worth-noticing. 1) The optimisation is constrained and the space of the bases is a $d$-dimensional sphere. 2) The objective function may not be differentiable for a constructed index function. 3) Although finding the global maximum is the goal of an optimisation problem, it is also interesting to inspect local maximum in projection pursuit since it could present unexpected interesting projections. 4) The optimisation procedure needs to be easy to compute since the tour animation needs to be played in real-time during the optimisation.

@@

\begin{algorithm}
\SetAlgoLined
\Repeat{$l > l_{\max}$ or $p_{\text{diff}} > 0.001$}{
  Generate $n$ random directions $\mathbf{A}_{\text{rand}}$ \;
  Compute $2n$ candidate bases deviate from $\mathbf{A}_{\text{cur}}$ by an angle of $\delta$ while ensuring orthogonality\;
  Compute the corresponding index value for each candidate bases\;
  Determine the search direction as from $\mathbf{A}_{\text{cur}}$ to the candidate bases with the largest index value\;
  Determine the step size via optimising the index value on the search direction over a 90 degree window\;
  find the optimum $\mathbf{A}_{**}$ and compute $I_{**} = f(\mathbf{A}_{**})$, $p_{\text{diff}} = (I_{**} - I_{\text{cur}})/I_{**}$\;
  update $l = l + 1$\;
}
\caption{Pseudo derivative (PD)}
\label{search-geodesic}
\end{algorithm}

@@

Three main pieces of information are recorded for the projection pursuit optimisers: 1) projection bases $\mathbf{A}$, 2) index values $I$, and 3) state $S$. For SA and SAJO, possible states include `random_search`,  `new_basis`, and `interpolation`. Pseudo derivative (PD) has a wider variety of states including `new_basis`, `direction_search`, `best_direction_search`, `best_line_search`, and `interpolation`. Multiple iterators index the information collected at different levels: $t$ is a unique identifier prescribing the natural ordering of each observation;  $j$ and $l$ are the counter of the outer and inner loop respectively. Other parameters of interest recorded include `method` that tags the name of the optimiser, and `alpha` that indicates the sampling neighbourhood size for searching observations.  A matrix notation of the data structure is presented in equation \ref{eq:data-structure}.

@@

\noindent where $T^{\prime} = T + k_{J}+ l_{J+1}$. Note that there is no output in iteration $J + 1$ since the optimiser does not find a better basis in the last iteration and terminates. The final basis found is $A_T$ with index value $I_T$.

@@

A starting point of diagnosing an optimiser is to understand how many searches an optimiser has conducted. One may want to simply plot the index value of the search points across its natural order, but a point geometry may work well if each iteration has a similar number of points. When some iterations have considerably more points than others, using a point geometry over-emphasizes the iterations that have more search points since these iterations will occupy the mast majority of the plot space. An alternative is to summarise the search in each iteration using boxplots and each iteration will then be spaced out equally. Occasionally, one may still want to switch back to a point geometry if the number of points is small in a particular iteration and this can be achieved via the `cutoff` argument in the search plot function. Additional annotations are added to facilitate better reading of the plot and these include  
1) the number of points searched in each iteration can be added as text label at the bottom of each iteration; 2) the anchor bases to interpolate are connected and highlighted in a larger size; and 3) the colour of the last iteration is in a grey scale to indicate no better basis found in this iteration.

@@

```{r toy-search, fig.cap="A comparison of search by two optimisers: SA (left) and SAJO (right) on a 2D projection of a six-variable dataset, \\code{boa6}, using the holes index. Both optimisers find the final basis with similar index value while it takes SAJO more iterations because the algorithm allows a probabilistic acceptance of bases with lower index value, as observed in iteration 4 and 6-8."}
```

@@

```{r toy-interp, fig.cap = "Trace plots of the interpolating bases with the same optimisation routine as the previous figure. Both traces are smooth while the change of index value in each iteration may not be monotonic."}
```

Another interesting diagnostic is to examine how the index value improves between interpolating bases since the projection on these bases is played by the tour animation. Trace plots are created with plotting the index value of the interpolation points against time. Figure \ref{fig:toy-interp} presents the trace plot of the same optimisation routine as Figure \ref{fig:toy-search}. The trace plot is smooth and an interesting discovery is that in SAJO, the interpolation in iteration 6-8 is not monotonic. Sometimes, this non-monotonicity is acceptable since the target basis does not necessarily have a higher index value than the current one, while in some cases, the optimisation will be better off if these intermediate basis can be used as the start of the next iteration (see section \ref{monotonic} for a detailed discussion on this).
